---
title: 'Intoducing ClustImpute: A new approach for k-means clustering with build-in
  missing data imputation'
author: Oliver Pfaffel
date: '2019-06-20'
categories:
  - R
  - statistical analysis
  - unsupervised machine learning
tags:
  - clustering
  - imputation
  - missing data
  - k-means
slug: introducing-clustimpute
comments: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, include = FALSE}
library(ClustImpute)
required_packages <- Hmisc::Cs(psych,ggplot2,tidyr,Hmisc,tictoc,ClusterR,copula,dplyr,corrplot)
lapply(required_packages, require, character.only = TRUE)
```

We are happily introducing a new k-means clustering algorithm that includes a powerful multiple missing data imputation at the computational cost of a few extra random imputations (benchmarks following in a separate article). More precisely, the algorithm draws the missing values iteratively based on the current cluster assignment so that correlations are considered on this level (we assume a more granular dependence structure is not relevant if we are "only" interest in k partitions). Subsequently, penalizing weights are imposed on imputed values and successivelydecreased (to zero) as the missing data imputation gets better. The hope is that at some point the observed point is near a cluster that provides a suitable neighborhood to draw the missing variable from. The algorithm is computationally efficient since the imputation is only as accurate as the clustering, and will be much faster than any approach that derives the full conditional missing distribution, e.g., as implemented in the (awesome) MICE package, independently of the clustering.

ClustImpute can currently be installed via github only:

```{r}
# devtools::install_github("o1iv3r/ClustImpute")
library(ClustImpute)
```

We'll provide an example based on simulated data to emphasis the benefits of ClustImpute.

# Simulated data with missings

First we create a random dataset with some structure and a few uncorrelated variables

```{r}
### Random Dataset
set.seed(739)
n <- 7500 # numer of points
nr_other_vars <- 4
mat <- matrix(rnorm(nr_other_vars*n),n,nr_other_vars)
me<-4 # mean
x <- c(rnorm(n/3,me/2,1),rnorm(2*n/3,-me/2,1)) 
y <- c(rnorm(n/3,0,1),rnorm(n/3,me,1),rnorm(n/3,-me,1))
true_clust <- c(rep(1,n/3),rep(2,n/3),rep(3,n/3)) # true clusters
dat <- cbind(mat,x,y)
dat<- as.data.frame(scale(dat)) # scaling
summary(dat)
```

One can clearly see the three clusters

```{r}
plot(dat$x,dat$y)
```

We create 20% of missings using the custom function **miss_sim()**

```{r}
dat_with_miss <- miss_sim(dat,p=.2,seed_nr=120)
summary(dat_with_miss)
```

The correlation matrix of the missing indicator shows that the missings are correlated - thus we are not in a missing completely at random (MCAR) stetting:

```{r}
mis_ind <- is.na(dat_with_miss) # missing indicator
corrplot(cor(mis_ind),method="number")
```


# Typical approach: median or random imputation

Clearly, an imputation with the median value does a pretty bad job here:

```{r}
dat_median_imp <- dat_with_miss
for (j in 1:dim(dat)[2]) {
  dat_median_imp[,j] <- Hmisc::impute(dat_median_imp[,j],fun=median)
}
imp <- factor(pmax(mis_ind[,5],mis_ind[,6]),labels=c("Original","Imputed")) # point is imputed if x or y is imputed
ggplot(dat_median_imp) + geom_point(aes(x=x,y=y,color=imp))
```

But also a random imputation is not much better: it creates plenty of points in areas with no data

```{r}
dat_random_imp <- dat_with_miss
for (j in 1:dim(dat)[2]) {
  dat_random_imp[,j] <- impute(dat_random_imp[,j],fun="random")
}
imp <- factor(pmax(mis_ind[,5],mis_ind[,6]),labels=c("Original","Imputed")) # point is imputed if x or y is imputed
ggplot(dat_random_imp) + geom_point(aes(x=x,y=y,color=imp))
```

A clustering based on random imputation will thus not provide good results (even if we "know" the number of clusters as in this example)

```{r}
tic("Clustering based on random imputation")
cl_compare <- KMeans_arma(data=dat_random_imp,clusters=3,n_iter=100,seed=751)
toc()
dat_random_imp$pred <- predict_KMeans(dat_random_imp,cl_compare)
ggplot(dat_random_imp) + geom_point(aes(x=x,y=y,color=factor(pred)))
```


# Better approach: ClustImpute

We'll now use ClustImpute and also measure the run-time. In short, the algorithm follows these steps

1. It replaces all NAs by random imputation, i.e., for each variable with missings, it draws from the marginal distribution of this variable not taking into account any correlations with other variables
2. Weights <1 are used to adjust the scale of an observation that was generated in step 1. The weights are calculated by a (linear) weight function that starts near zero and converges to 1 at n_end.
3. A k-means clustering is performed with a number of c_steps steps starting with a random initialization.
4. The values from step 2 are replaced by new draws conditionally on the assigned cluster from step 3.
5. Steps 2-4 are repeated nr_iter times in total. The k-means clustering in step 3 uses the previous cluster centroids for initialization.
6. After the last draws a final k-means clustering is performed.

The intuition is that points should be clustered with other points mainly based on their observed values, while the resulting clusters provide donors for the missing value imputation, so that step by step all variables can be used for the clustering. 


```{r}
nr_iter <- 10 # iterations of procedure
n_end <- 10 # step until convergence of weight function to 1
nr_cluster <- 3 # number of clusters
c_steps <- 50 # numer of cluster steps per iteration
tic("Run ClustImpute")
res <- ClustImpute(dat_with_miss,nr_cluster=nr_cluster, nr_iter=nr_iter, c_steps=c_steps, n_end=n_end) 
toc()
```

ClustImpute provides several results:

```{r}
str(res)
```

We'll first look at the complete data and clustering results. Quite obviously, it gives better results then median / random imputation.

```{r}
ggplot(res$complete_data,aes(x,y,color=factor(res$clusters))) + geom_point()
```

Packages like MICE compute a traceplot of mean and variance of the imputed variables for various chains. This diagnostics helps to show if the Markov chains converge to a stationary distribution. 

Here we only have a single realization  and thus re-run ClustImpute with various seeds to obtain different realizations. 

```{r}
res2 <- ClustImpute(dat_with_miss,nr_cluster=nr_cluster, nr_iter=nr_iter, c_steps=c_steps, n_end=n_end,seed_nr = 2)
res3 <- ClustImpute(dat_with_miss,nr_cluster=nr_cluster, nr_iter=nr_iter, c_steps=c_steps, n_end=n_end,seed_nr = 3)
mean_all <- rbind(res$imp_values_mean,res2$imp_values_mean,res3$imp_values_mean)
sd_all <- rbind(res$imp_values_sd,res2$imp_values_sd,res3$imp_values_sd)

mean_all <- cbind(mean_all,seed=rep(c(150519,2,3),each=11))
sd_all <- cbind(sd_all,seed=rep(c(150519,2,3),each=11))
```

The realizations mix nicely with each other, as the following plots shows. Thus it seems we obtain to a similar missing value distribution independently of the seed.

```{r}
ggplot(as.data.frame(mean_all)) + geom_line(aes(x=iter,y=V1,color=factor(seed))) + ggtitle("Mean")
ggplot(as.data.frame(sd_all)) + geom_line(aes(x=iter,y=V1,color=factor(seed))) + ggtitle("Std. dev.")
```

# Quality of imputation and cluster results

## Marginal distributions

Now we compare the marginal distributions using a violin plot of x and y. In particular for y, the distribution by cluster is quite far away from the original distribution for the random imputation based clustering, but quite close for ClustImpute.

```{r}
dat4plot <- dat
dat4plot$true_clust <- true_clust
Xfinal <- res$complete_data
Xfinal$pred <- res$clusters

par(mfrow=c(1,2))
violinBy(dat4plot,"x","true_clust",main="Original data")
violinBy(dat4plot,"y","true_clust",main="Original data")
violinBy(Xfinal,"x","pred",main="imputed data")
violinBy(Xfinal,"y","pred",main="imputed data")
violinBy(dat_random_imp,"x","pred",main="random imputation")
violinBy(dat_random_imp,"y","pred",main="random imputation")
```

## External validation: rand index

Below we compare the rand index between true and fitted cluster assignment. For ClustImpute we obtain:

```{r}
external_validation(true_clust, res$clusters)
```

This is a much higher value than for a clsutering based on random imputation:

```{r}
class(dat_random_imp$pred) <- "numeric"
external_validation(true_clust, dat_random_imp$pred)
```

Not surprisingly, the RandIndex for ClustImpute is much higher if we consider complete cases only (and throw away a considerable amount of our data).

```{r}
## complete cases
idx <- which(complete.cases(dat_with_miss)==TRUE)
sprintf("Number of complete cases is %s",length(idx))
sprintf("Rand index for this case %s", external_validation(true_clust[idx], res$clusters[idx]))
```

Aside from the RandIndex, this function also computes a variety of other stats

```{r}
external_validation(true_clust, res$clusters,summary_stats = TRUE)
```


## Variance reduction

To assess quality of our cluster results, we compute the sum of squares within each cluster, sum up these values and compare it with the total sum of squares.

```{r}
res_var <- var_reduction(res)
res_var$Variance_reduction
res_var$Variance_by_cluster
```

We se a reduction of about 27\% using only 3 clusters, most strikingly for x and y because that these variables define the subspace of the true clusters.

More clusters will capture the random distribution of the other variables

```{r}
res <- ClustImpute(dat_with_miss,nr_cluster=10, nr_iter=nr_iter, c_steps=c_steps, n_end=n_end)
res_var <- var_reduction(res)
res_var$Variance_reduction
res_var$Variance_by_cluster
```

Can we use this function systematically to find out the optimal number of clusters? We'll do the exercise above for a several values of nr_cluster and need a helper function for that since X is an argument of ClustImpute

```{r}
ClustImpute2 <- function(dataFrame,nr_cluster, nr_iter=10, c_steps=1, wf=default_wf, n_end=10, seed_nr=150519) {
  return(ClustImpute(dataFrame,nr_cluster, nr_iter, c_steps, wf, n_end, seed_nr))
}
res_list <- lapply(X=1:10,FUN=ClustImpute2,dataFrame=dat_with_miss, nr_iter=nr_iter, c_steps=c_steps, n_end=n_end)
```

Nex we put the variances by cluster in a table

```{r}
tmp <- var_reduction(res_list[[1]])
var_by_clust <- tmp$Variance_by_cluster
for (k in 2:10) {
  tmp <- var_reduction(res_list[[k]])
  var_by_clust <- rbind(var_by_clust,tmp$Variance_by_cluster)
}
var_by_clust$nr_clusters <- 1:10
```


While there is a rather gradual improvement for the other variables, x and y have a minimum at 3 showing optimality for these variables. Such a plot clearly indicates that 3 clusters are a good choice for this data set (which, of course, we knew in advance)

```{r}
data2plot <- tidyr::gather(var_by_clust,key = "variable", value = "variance", -dplyr::one_of("nr_clusters"))
ggplot(data2plot,aes(x=nr_clusters,y=variance,color=variable)) + geom_line() + scale_x_continuous(breaks=1:10)
```

Now it's time for you to try this algorithm on real problems! Looking forward to feedback  via twitter or on the github page.

